---
title: "Gaussian Plume Model"
description: "Simple steady state model for instaneous emission distribution."
date: "2025-04-13"
categories: [R, HPAI]
format:
  gfm:
   toc: true
   toc-depth: 2
   code-fold: show
   code-summary: "Hide code"
   preview-mode: raw
editor: source
editor_options:  
  chunk_output_type: console
---

## Libraries
```{r warning=FALSE, messages=FALSE}
library(here)
library(tidyverse)
library(ggmap)
library(ggspatial)
library(sf)
library(terra)
```


## Custom Functions
```{r}
source(here("R/utilities.R"))
source_dir(here("R"))
```

## Create Study Area Grid  
Choosing an arbitrary geographic point.  It will serve as the emission point source location and be used to define study area extent.  
```{r}
center_coordinate <- c(-88.7904, 33.4555)  # Miss State Univ, center of study area
source_vect <- vect(matrix(center_coordinate, ncol = 2), type = "points", crs = "EPSG:4326")
timezone_utc <- get_timezone_utc(center_coordinate); timezone_utc

# desired projection
user_projection <- "+proj=utm +zone=18 +datum=WGS84 +units=m +no_defs"

# create grids
grid_raster <- create_spatraster_grid(center_coordinate, user_projection, extent_km = 25) # projected
grid_raster_geo <- create_spatraster_grid_geo(center_coordinate) # no proj version

# project source/center point
source_vect <- project(source_vect, user_projection)
```


## Location Map 
Register Stadia Maps API to pull background images.
```{r}
# register my API
map_api <- yaml::read_yaml(here("local", "secrets.yaml"))
register_stadiamaps(key = map_api$stadi_api)
```

```{r fig.width=10, fig.height=10}
map_plot <- map_grid(grid_raster)
map_plot
```


## Gaussian Plume
The Gaussian plume model in by *estimate_gaussian_plume()*  uses the standard equation for continuous emissions from a point source in a steady-state wind field, assuming Gaussian distribution of concentrations in the horizontal and vertical directions.  
  
**Math notation:**    
 
\[
C(x, y, z) = \frac{Q}{2\pi u \sigma_y \sigma_z} \exp\left(-\frac{y^2}{2\sigma_y^2}\right) \left[ \exp\left(-\frac{(z - H)^2}{2\sigma_z^2}\right) + \exp\left(-\frac{(z + H)^2}{2\sigma_z^2}\right) \right]
\]
  
**Where:**
- \(C(x, y, z)\) is the concentration at point \((x, y, z)\),
- \(Q\) is the emission rate (g/s),
- \(u\) is the wind speed (m/s),
- \(\sigma_y(x)\) and \(\sigma_z(x)\) are the standard deviations of the concentration distribution in the lateral and vertical directions, respectively (based on downwind distance \(x\) and atmospheric stability class),
- \(H\) is the effective stack height (m),
- \(x\) is the downwind distance,
- \(y\) is the crosswind distance,
- \(z\) is the height above ground.

The \(y\) and \(z\) bits correspond to `f_y` and `g1`, `g2` in *estimate_gaussian_plume()*, see the [code HERE]().



### Organize Data  
Organizing location data to predict and visualize particle concentrations across study area.  
```{r}
# source farm
source_coords <- crds(source_vect)

# cell coordinates
all_cells <- 1:ncell(grid_raster)
xy <- as.data.frame(xyFromCell(grid_raster, all_cells))

# relative coordinates with respect to source (assuming wind from west to east)
xy$x_rel <- xy$x - source_coords[1]
xy$y_rel <- xy$y - source_coords[2]

# assume all receptors are at 1.8 m height
receptors <- data.frame(x = xy$x_rel, y = xy$y_rel, z = rep(1.8, nrow(xy)))
```


## Plume Parameters  
```{r}
# poultry house dimensions 13m x 152 m = 1976 x 3m tall = 5928 m3
# average exhaust 3.10 m3/h
# PM_2.5 0.149 mg/m3

Q <- 100   # Emission rate (g/s)
H <- 10   # Source height (m)
u <- 3    # Wind speed (m/s)
```

## Pasquill Stability Classes  
See, [NOAA page HERE](https://www.ready.noaa.gov/READYpgclass.php)  
   
- A, B, and C are considered unstable conditions, with A being very unstable, producing high dispersion due to strong vertical mixing).  
- D represents neutral conditions—typically found under overcast skies or in urban areas where there is little convective activity.  
- E and F are stable conditions, with F being the most stable, with little vertical mixing (plume remains more confined).   
   
## Create Plume Function 
```{r}
plume_model <- estimate_gaussian_plume(Q, H, u, sigma = stability_class("F"))
```

## Apply Plume Function
```{r}
downwind_idx <- receptors$x > 0 # behind source (assuming wind W -> E)

# concentrations vector
concentrations <- numeric(nrow(receptors))
concentrations[!downwind_idx] <- 0
concentrations[downwind_idx] <- plume_model(receptors[downwind_idx, ])

# assign concentrations to grid
concentration_raster <- setValues(grid_raster, concentrations)
```


## Plot Results
Convert to dataframea.  Setting zero concentrations to NA
```{r}
df <- as.data.frame(concentration_raster, xy = TRUE)
df$grid[df$grid == 0] = NA

source_df <- data.frame(x = source_coords[,1], y = source_coords[,2])
```


Basic grid view
```{r fig.width=10, fig.height=10}
ggplot() +
  geom_raster(data = df, aes(x = x, y = y, fill = grid)) +
  scale_fill_gradientn(
    colors = rev(pals::cubehelix(30)[1:26]),
    na.value = "white",
    limits = c(0, max(df$grid, na.rm = TRUE)),
    name = "Concentration (g/m3)"
  ) +
  geom_point(data = source_df,
             aes(x = x, y = y),
             color = "blue", size = 3) +
  coord_equal() +
  labs(title = "Plume Dispersion",
       x = "Easting (m)",
       y = "Northing (m)") +
  theme_minimal()
```

Overlay on map
```{r fig.width=10, fig.height=10}
map_plot_plume <- map_grid(grid_raster, 
                           concentration_raster = concentration_raster, 
                           map_type = "stamen_toner_lite")
map_plot_plume
```